Learning Causally-Generated Stationary Time Series

### 2017-02-24 doc/paper.tex
First draft

\begin{abstract}
We investigate the notion of causality in modelling stationary time series. Extending on the work by \citet{Tobar:2015:Learning_Stationary}, we present the Causal Gaussian Process Convolution Model (CGPCM). The CGPCM models the kernel of a Gaussian process nonparametrically and has an inductive bias towards causally-generated time series. We develop an algorithm to perform inference and apply the CGPCM to tasks involving real-world data.
\end{abstract}

@@

One of the major goals of statistical inference is to develop models of dynamical phenomena that can be used for prediction, extrapolation, and system identification. Two key characteristics of the physical systems underlying natural and manmade dynamical phenomena are that they are \textit{causal} and \textit{spectrally complex}. Causal systems, where at any point in time the output of the system can only depend on past values of the input, are the only type which are physically realisable. Spectrally complex systems, which have rich power spectral densities, arise because many physical systems have numerous resonances spanning many time scales. When developing statistical models for such phenomena this prior knowledge---causality and spectral richness---should be leveraged in order to exclude any unrealisable system from the model prior whilst at the same time allowing the model to have the capacity to capture varied spectral content that may be only slowly revealed as more data are seen. The goal of this paper is to develop Gaussian Process (GP) models together with associated inference and learning schemes that serve this purpose.

A widely-used model for stationary time series are Gaussian Processes, which place a prior distribution over the latent function underlying the time series $f:\R \to \R$ by assuming that any finite collection of function values $f(t_1),\ldots,f(t_n)$ is multivariate-Gaussian distributed. The key modelling decision in using a Gaussian Process is the choice of covariance $k_f(t,t')$ between any two function values $f(t)$ and $f(t')$---$k_f$ is also called the \textit{kernel} of $f$. This choice of kernel\footnote{The mean of the function is assumed to be zero without loss of generality. \cite{Rasmussen:2006:Gaussian_Processes}} encodes prior information about the function $f$. Consequently, a large research effort has been devoted towards developing flexible kernels \cite{Duvenaud:2014:Automatic_Construction,Wilson:2013:Spectral_Mixture,Tobar:2015:Learning_Stationary,Tobar:2015:Inter-Domain_Inducing}.

Of particular interest for the current application are GP models that use nonparametric models of the power spectra of the GPs---doubly non-parametric models---as they have the capacity to flexibly model signals with arbitrary spectral complexity. One approach employs a Dirichlet Process Mixture Model for the power spectra \cite{Oliva:2015:Bayesian_Nonparametric_Kernel-Learning}, but it is not clear whether it is possible to build the causality constraint into such a construction. An alternative approach instead induces a nonparametric power spectrum by placing a Gaussian Process prior over the system's impulse response function \cite{Tobar:2015:Learning_Stationary}. Critically, the so-called Gaussian Process Convolution Model (GPCM), does not incorporate a causality constraint---the impulse response in question should be thought of as an acausal spatial response. In this paper we revisit the GPCM and explicitly build in a causality constraint whilst retaining the flexibility of modelling the kernel nonparametrically. In addition to these modelling contributions, we develop enhanced variational inference and learning schemes for the causal and acausal variants of the GPCM---collapsed variational bounds and structured approximating distributions---that significantly improve statistical accuracy. These modelling and inferential contributions are demonstrated on synthetic and real-world signals.

@@

One of the focusses in this work is to develop flexible prior distributions over the filter $h$ that can support essentially arbitrarily complex structure. We therefore choose to model it using a Gaussian process $h \sim \mathcal{GP}(0,k_h(t,t'))$. The covariance function describing the prior over the filter$k_h(t,t')$ should be carefully chosen. One important constraint on the covariance function comes from the fact that every real-world signal $f$ has finite power. This can be satisfied by letting the filter variance decay to zero at infinity sufficiently quickly \citet{Tobar:2015:Learning_Stationary}: let $h\cond g = w g$ where $w(t)= \exp(- \alpha t^2)$ and $g \sim \mathcal{GP}(0,k_g(t-t'))$ has finite power; in that case, % RET there's a question as to whether exponential decay might be more appropriate in the current case
\begin{align*}
    \mathbb{V}[f(t)]
    &= \int_{0}\exp(-2 \alpha t^2)\mathbb{V}[g(\tau)]\id{\tau} \\
    &\le k_g(0) \int_0 \exp(- 2 \alpha t^2) \id{\tau} < \infty,
\end{align*}
which could be infinite otherwise.
Equivalently, we let $k_h(t,t')=\exp(- \alpha (t^2 + t^{\prime 2}))k_g(t-t')$. To retain flexibility in the prior on $h$, we let $k_g$ be an exponentiated quadratic. We thus have that $k_h(t,t')=\exp(- \alpha (t^2 + t^{\prime 2}) - \gamma(t-t')^2)$. Note that $\alpha$ determines the typical temporal extent of the filter and $\gamma$ the typical time-scale over which it varies. These assumptions also serve to make inference well-posed; without them, any shifted version of the filter results in an identical statistical model for the data meaning the model is unidentifiable.

@@

\Cref{mod:cgpcm2} reveals the CGCPM as a Gaussian process in which the kernel, or equivalently the power spectral density (PSD), is modelled nonparametrically. \Cref{fig:interpolation} illustrates the generative process of the CGPCM: First, a filter $h$ is generated. Then, the kernel $k_{f\cond h}$ is constructed. Finally, a sample $f\cond h$ is drawn from $\mathcal{GP}(0,k_{f \cond h}(t-t'))$. \Cref{fig:prior} visualises the corresponding prior over kernels, and \cref{fig:prior_psd} visualises the corresponding prior over PSDs. Note that the prior over PSDs for the GPCM has a confident low-pass structure, whereas that for the CGPCM has support for more slowly decaying spectra.

@@

\begin{figure}[t]
    \centering
    \begin{tabularx}{\linewidth}{>{\centering}X>{\centering}X}
        GPCM & CGPCM
    \end{tabularx}
    \includegraphics[width=\linewidth, height=3cm]{resources/cropped/prior.pdf}
    \caption{Prior distribution over kernels in the GPCM and CGPCM. Lines correspond to means and gradients correspond to uncertainty.}
    \label{fig:prior}
\end{figure}

\begin{figure}[t]
    \centering
    \begin{tabularx}{\linewidth}{>{\centering}X>{\centering}X}
        ~~GPCM & ~~~~~~CGPCM
    \end{tabularx}
    \includegraphics[width=\linewidth, height=3cm]{resources/cropped/prior_psd.pdf}
    \caption{Prior distribution over PSDs in the GPCM and CGPCM. The unit on the $y$ axis is decibel. Lines correspond to means and gradients correspond to uncertainty.}
    \label{fig:prior_psd}
\end{figure}

\subsection{Sampling from the CGPCM}
Sampling from the CGPCM is challenging because the integrals in \cref{mod:cgpcm,mod:cgpcm2} depend on the entirety of $h$ and $x$. To resolve this issue, we follow \citet{Tobar:2015:Learning_Stationary} and approximate \cref{eq:kernel} using Bayesian quadrature \cite{Minka:2000:Quadrature_GP} by conditioning on finitely many values $u=(h(t_{u,1}),\ldots,h(t_{u,n_u}))\sim \mathcal{N}(0,K_u)$ of $h$, which induces a  distribution over covariance functions. The mean of this distribution is an accurate approximation for $k_{f\cond h}$ if sufficient conditioning values are used,
\begin{align*}
    k_{f\cond h}(r)
    &\approx \mathbb{E}[k_{f\cond h}(r)\cond u]
    = \int_{0} \mathbb{E}[h(|r| + \tau) h(\tau)\cond u] \id{\tau} \\
    &= \int_0 k_h(|r| + \tau, \tau) \id{\tau} \\
    &\phan{=}+ \mathrm{trace} \left( M^u \int_{0} k_h(t_u, |r|+ \tau)k_h(\tau, t_u^T) \id{\tau} \right)
\end{align*}
where $M^{u}=K_u^{-1}uu^T K_u^{-1}-K_u$. This expression was used to compute the model samples shown in \cref{fig:interpolation}. %Unlike the GPCM, the kernel approximation $\mathbb{E}[k_{f\cond h}(r)\cond u]$ of the CGPCM will not simply be a mixture of Gaussians; it will be a more complicated expression instead.

@@

Following an argument similar, we again derive the corresponding ELBO:
\begin{align*}
    &\mathcal{L}[q(u),q(z\cond u)] \\
    &\quad= \la \log p(e\cond f) \ra_{q(f,u,z)}- \ll\la\log\frac{q(u)q(z\cond u)}{p(u)p(z)}\rr\ra_{q(u)q(z\cond u)}.
\end{align*}
Again, setting the variations $\delta \mathcal{L} / \delta q(u)$ and $\delta \mathcal{L} / \delta q(z\cond u)$ to zero and solving for for respectively $q(u)$ and $q(z\cond u)$ yields that
\begin{align}
    q(u) &\propto p(u) \int p(z) \exp\la\log p(e\cond f)\ra_{p(f\cond u, z)}\id{z}, \label{eq:qu-smf} \\
    q(z\cond u) &\propto p(z)\exp\la \log p(e\cond f)\ra_{p(f\cond u, z)}, \label{eq:qz-smf}
\end{align}
which are computed in \cref{app:computation_quz}.
As opposed to \cref{eq:qu,eq:qz}, \cref{eq:qu-smf,eq:qz-smf} are uncoupled, but $q(u)$'s moments are now intractable. We can, however, evaluate $q(u)$. We therefore use elliptical slice sampling (ESS) \cite{Murray:2010:Elliptical_Slice_Sampling} to sample from $q(u)$ and use these samples to approximate $q(f, u, z)$. To help mixing the Markov chain, we initialise the sampler with the MF approximation.

\section{Experiments}
In this section we test the CGPCM by applying it to synthetic and real-world signals. We show that in certain situations the causality constraint provides a useful inductive bias that leads to better predictions.

@@

\begin{figure*}[t]
    \centering
    GPCM
    \begin{tabularx}{\linewidth}{
            >{\centering}>{\hsize=.43\hsize}X  % ??
            >{\centering}>{\hsize=.2\hsize}X
            >{\centering}>{\hsize=.2\hsize}X
            >{\centering}>{\hsize=.2\hsize}X
        }
        $f\cond h$ & $h$ & $k_{f\cond h}$ & PSD of $f\cond h$
    \end{tabularx}
    \includegraphics[width=\linewidth, height=3cm]{resources/cropped/learning_known_kernels_acausal_sample_gpcm.pdf}
    CGPCM
    \begin{tabularx}{\linewidth}{
            >{\centering}>{\hsize=.43\hsize}X  % ??
            >{\centering}>{\hsize=.2\hsize}X
            >{\centering}>{\hsize=.2\hsize}X
            >{\centering}>{\hsize=.2\hsize}X
        }
        $f\cond h$ & $h$ & $k_{f\cond h}$ & PSD of $f\cond h$
    \end{tabularx}
    \includegraphics[width=\linewidth, height=3cm]{resources/cropped/learning_known_kernels_acausal_sample_cgpcm.pdf}
    \caption{Predictions of the GPCM and CGPCM after learning a noisy random sample from the GPCM. This sample consists of 300 data points. We used 51 evenly-spaced inducing points for $h$ and 80 evenly-spaced inducing points for $x$; inducing points are indicated by black dots in the figure. All parameters were initialised randomly and the procedure from \cref{sec:inference} was followed. All signals are normalised to unity power. Finally, {\color{red}red} denotes truth; {\color{darkgreen}green} denotes observations, the emperical autocorrelation, and the periodogram for respectively the function $f\cond h$, the kernel $k_{f\cond h}$, and the PSD of $f\cond h$; {\color{blue}blue} denotes quantities learned through the MF approximation; and {\color{mycyan}cyan} denotes quantities learned through the SMF approximation. For any learned quantity it holds that lines correspond to means and areas correpond to 95\% central credible regions.}
    \label{fig:toy_acausal_sample}
\end{figure*}

@@

\begin{figure*}[t]
    \centering
    GPCM
    \begin{tabularx}{\linewidth}{
            >{\centering}>{\hsize=.43\hsize}X  % ??
            >{\centering}>{\hsize=.2\hsize}X
            >{\centering}>{\hsize=.2\hsize}X
            >{\centering}>{\hsize=.2\hsize}X
        }
        $f\cond h$ & $h$ & $k_{f\cond h}$ & PSD of $f\cond h$
    \end{tabularx}
    \includegraphics[width=\linewidth, height=3cm]{resources/cropped/learning_known_kernels_causal_sample_gpcm.pdf}
    CGPCM
    \begin{tabularx}{\linewidth}{
            >{\centering}>{\hsize=.43\hsize}X  % ??
            >{\centering}>{\hsize=.2\hsize}X
            >{\centering}>{\hsize=.2\hsize}X
            >{\centering}>{\hsize=.2\hsize}X
        }
        $f\cond h$ & $h$ & $k_{f\cond h}$ & PSD of $f\cond h$
    \end{tabularx}
    \includegraphics[width=\linewidth, height=3cm]{resources/cropped/learning_known_kernels_causal_sample_cgpcm.pdf}
    \caption{Predictions of the GPCM and CGPCM after learning a noiseless random sample from the CGPCM. The sample consists of 300 data points. We used 51 evenly-spaced inducing points for $h$ and 80 evenly-spaced inducing points for $x$; inducing points are indicated by black dots in the figure. All parameters were initialised randomly and the procedure from \cref{sec:inference} was followed. All signals are normalised to unity power. Finally, {\color{red}red} denotes truth; {\color{darkgreen}green} denotes observations, the emperical autocorrelation, and the periodogram for respectively the function $f\cond h$, the kernel $k_{f\cond h}$, and the PSD of $f\cond h$; {\color{blue}blue} denotes quantities learned through the MF approximation; and {\color{mycyan}cyan} denotes quantities learned through the SMF approximation. For any learned quantity it holds that lines correspond to means and areas correpond to 95\% central credible regions.}
    \label{fig:toy_causal_sample}
\end{figure*}

@@

Finally, note that predictions through the SMF approximation consistently give better-calibrated credible regions than predictions through the MF approximation. We will further investigate issue in the following experiment.

\subsection{Speech}
Coming up.

@@

\subsection{Head-Related Impulse Response Estimation}
The head-related impulse response (HRIR) describes how the reception of sound by the ear depends on the position from which the sound originates. Specifically, the received sound is convolved with the HRIR before it is sensed by the ear, and the HRIR depends on the direction from which the sound comes. We consider the problem of inferring the HRIR after some sound has been sensed. Suppose that the sound is background noise and can be modelled by white noise. Then assuming a Gaussian process prior over the HRIR recovers exactly the CGPCM; therefore, in that case, we can infer the HRIR by inferring the filter $h$ in the CGPCM.

\begin{figure*}[t]
    \centering
    \begin{tabularx}{\linewidth}{>{\centering}X>{\centering}X}
        GPCM & CGPCM
    \end{tabularx}
    \begin{tabularx}{\linewidth}{>{\centering}X>{\centering}X>{\centering}X>{\centering}X}
        $k_{f\cond h}$ & $h$ & $k_{f\cond h}$ & $h$
    \end{tabularx}
    \includegraphics[width=.49\linewidth, height=3cm]{resources/hrtf_gpcm.pdf}
    \includegraphics[width=.49\linewidth, height=3cm]{resources/hrtf_cgpcm.pdf}
    \caption{Predictions of the GPCM and CGPCM after learning a HRIR convolved with white noise. The sample consists of 500 data points. We used 101 evenly-spaced inducing points for $h$ and 150 evenly-spaced inducing points for $x$; inducing points are indicated by black dots in the figure. All parameters were initialised randomly and the procedure from \cref{sec:inference} was followed. All signals are normalised to unity power. Here {\color{red}red} denotes truth, {\color{blue}blue} denotes quantities learned through the MF approximation, and {\color{mycyan}cyan} denotes quantities learned through the SMF approximation. For any learned quantity it holds that lines correspond to means and areas correpond to 95\% central credible regions.}
    \label{fig:hrir}
\end{figure*}

@@

\section{Discussion}
We have presented the CGPCM as a model of causal, spectrally complex dynamical phenomena. Inference in the CGPCM is performed in a two-step procedure: first a collapsed varational MF approach is used to obtain an initial approximation, and then this approximation is refined through a variational SMF approach combined with ESS. The CGPCM has been tested and validated on synthesised and real-world signals and shows encouraging results. In particular, the CGPCM shows the capability of modelling rough signals without attributing this roughness to observation noise.

>>>>

\begin{abstract}
We investigate the notion of causality in modelling stationary time series. Extending on the work by \citet{Tobar:2015:Learning_Stationary}, we present the Causal Gaussian Process Convolution Model (CGPCM). The CGPCM models the kernel of a Gaussian process nonparametrically and has an inductive bias towards causally-generated time series. We develop an algorithm to perform inference and demonstrate the CGPCM on synthetic and real-world signals.
\end{abstract}

@@

One of the major goals of statistical inference is to develop models of dynamical phenomena that can be used for prediction, extrapolation, and system identification. Two key characteristics of the physical systems underlying natural and manmade dynamical phenomena are that they are \textit{causal} and \textit{spectrally complex}. Causal systems, where at any point in time the output of the system can only depend on past values of the input, are the only type which are physically realisable. Spectrally complex systems, which have rich power spectral densities, arise because many physical systems have numerous resonances spanning many time scales. When developing statistical models for such phenomena this prior knowledge---causality and spectral richness---should be leveraged in order to exclude any unrealisable system from the model prior whilst at the same time allowing the model to have the capacity to capture varied spectral content that may be only slowly revealed as more data are seen. The goal of this paper is to develop Gaussian process (GP) models together with associated inference and learning schemes that serve this purpose.

Gaussian processes are a widely-used model for stationary time series. They place a prior distribution over the latent function underlying the time series $f:\R \to \R$ by assuming that any finite collection of function values $f(t_1),\ldots,f(t_n)$ is multivariate-Gaussian distributed. The key modelling decision in using a Gaussian process is the choice of covariance $k_f(t,t')$ between any two function values $f(t)$ and $f(t')$---$k_f$ is also called the \textit{kernel} of $f$. This choice of kernel encodes prior information about the function $f$. Consequently, a large research effort has been devoted towards developing flexible kernels \cite{Duvenaud:2014:Automatic_Construction,Wilson:2013:Spectral_Mixture,Tobar:2015:Learning_Stationary,Tobar:2015:Inter-Domain_Inducing}.

Of particular interest for the current application are GP models that use nonparametric models of the power spectra of the GPs---doubly nonparametric models---as they have the capacity to flexibly model signals with arbitrary spectral complexity. One approach employs a Dirichlet Process Mixture Model for the power spectra \cite{Oliva:2015:Bayesian_Nonparametric_Kernel-Learning}, but it is not clear whether it is possible to build the causality constraint into such a construction. An alternative approach instead induces a nonparametric power spectrum by placing a Gaussian process prior over the system's impulse response function \cite{Tobar:2015:Learning_Stationary}. Critically, the so-called Gaussian Process Convolution Model (GPCM), does not incorporate a causality constraint---the impulse response in question should be thought of as an acausal spatial response. In this paper we revisit the GPCM and explicitly build in a causality constraint whilst retaining the flexibility of modelling the kernel nonparametrically. In addition to these modelling contributions, we develop enhanced variational inference and learning schemes for the causal and acausal variants of the GPCM---collapsed variational bounds and structured approximating distributions---that significantly improve statistical accuracy. These modelling and inferential contributions are demonstrated on synthetic and real-world signals.

@@

One of the focusses in this work is to develop flexible prior distributions over the filter $h$ that can support essentially arbitrarily complex structure. We therefore choose to model it using a Gaussian process $h \sim \mathcal{GP}(0,k_h(t,t'))$. The covariance function describing the prior over the filter $k_h(t,t')$ should be carefully chosen. One important constraint on the covariance function comes from the fact that every real-world signal $f$ has finite power. This can be satisfied by letting the filter variance decay to zero at infinity sufficiently quickly \cite{Tobar:2015:Learning_Stationary}: let $h\cond g = w g$ where $w(t)= \exp(- \alpha t^2)$ and $g \sim \mathcal{GP}(0,k_g(t-t'))$ has finite power; in that case, % RET there's a question as to whether exponential decay might be more appropriate in the current case
\begin{align*}
    \mathbb{V}[f(t)]
    &= \int_{0}\exp(-2 \alpha t^2)\mathbb{V}[g(\tau)]\id{\tau} \\
    &\le k_g(0) \int_0 \exp(- 2 \alpha t^2) \id{\tau} < \infty,
\end{align*}
which could be infinite otherwise.
Equivalently, we let $k_h(t,t')=\exp(- \alpha (t^2 + t^{\prime 2}))k_g(t-t')$. To retain flexibility in the prior on $h$, we let $k_g$ be an exponentiated quadratic. We thus have that $k_h(t,t')=\exp(- \alpha (t^2 + t^{\prime 2}) - \gamma(t-t')^2)$. Note that $\alpha$ determines the typical temporal extent of the filter and $\gamma$ the typical time-scale over which it varies. These assumptions also serve to make inference well-posed; without them, any shifted version of the filter results in an identical statistical model for the data meaning the model is unidentifiable.

@@

\Cref{mod:cgpcm2} reveals the CGCPM as a Gaussian process in which the kernel, or equivalently the power spectral density (PSD), is modelled nonparametrically. \Cref{fig:interpolation} illustrates the generative process of the CGPCM: First, a filter $h$ is generated. Then, the kernel $k_{f\cond h}$ is constructed. Finally, a sample $f\cond h$ is drawn from $\mathcal{GP}(0,k_{f \cond h}(t-t'))$. \Cref{fig:prior} visualises the corresponding prior over kernels, and \cref{fig:prior_psd} visualises the corresponding prior over PSDs. Note that the prior over PSDs for the GPCM confidently has a low-pass structure, whereas that for the CGPCM has support for more slowly decaying spectra.

@@

\begin{figure}[t]
    \centering
    \begin{tabularx}{\linewidth}{>{\centering}X>{\centering}X}
        GPCM & CGPCM
    \end{tabularx}
    \includegraphics[width=\linewidth, height=3cm]{resources/cropped/prior.pdf}
    \caption{Prior distribution over kernels in the GPCM and CGPCM. Lines correspond to means and gradients correspond to uncertainty.}
    \label{fig:prior}
\end{figure}
\begin{figure}[t]
    \centering
    \begin{tabularx}{\linewidth}{>{\centering}X>{\centering}X}
        ~~GPCM & ~~~~~~CGPCM
    \end{tabularx}
    \includegraphics[width=\linewidth, height=3cm]{resources/cropped/prior_psd.pdf}
    \caption{Prior distribution over PSDs in the GPCM and CGPCM. The unit on the $y$ axis is decibel. Lines correspond to means and gradients correspond to uncertainty.}
    \label{fig:prior_psd}
\end{figure}

\subsection{Sampling from the CGPCM}
Sampling from the CGPCM is challenging because the integrals in \cref{mod:cgpcm,mod:cgpcm2} depend on the entirety of $h$ and $x$. To resolve this issue, we follow \citet{Tobar:2015:Learning_Stationary} and approximate \cref{eq:kernel} using Bayesian quadrature \cite{Minka:2000:Quadrature_GP} by conditioning on finitely many values $u=(h(t_{u,1}),\ldots,h(t_{u,n_u}))\sim \mathcal{N}(0,K_u)$ of $h$, which induces a  distribution over covariance functions. The mean of this distribution is an accurate approximation for $k_{f\cond h}$ if sufficient conditioning values are used,
\begin{align*}
    k_{f\cond h}(r)
    &\approx \mathbb{E}[k_{f\cond h}(r)\cond u]
    = \int_{0} \mathbb{E}[h(|r| + \tau) h(\tau)\cond u] \id{\tau} \\
    &= \int_0 k_h(|r| + \tau, \tau) \id{\tau} \\
    &\phan{=}+ \mathrm{trace} \left( M^u \int_{0} k_h(t_u, |r|+ \tau)k_h(\tau, t_u^T) \id{\tau} \right)
\end{align*}
where $M^{u}=K_u^{-1}uu^T K_u^{-1}-K_u$. This expression was used to compute the model samples shown in \cref{fig:prior,fig:prior_psd,fig:interpolation}. %Unlike the GPCM, the kernel approximation $\mathbb{E}[k_{f\cond h}(r)\cond u]$ of the CGPCM will not simply be a mixture of Gaussians; it will be a more complicated expression instead.

@@

Following an argument similar, we again derive the corresponding ELBO:
\begin{align*}
    &\mathcal{L}[q(u),q(z\cond u)] \\
    &\quad= \la \log p(e\cond f) \ra_{q(f,u,z)}- \ll\la\log\frac{q(u)q(z\cond u)}{p(u)p(z)}\rr\ra_{q(u)q(z\cond u)}.
\end{align*}
Again, setting the variations $\delta \mathcal{L} / \delta q(u)$ and $\delta \mathcal{L} / \delta q(z\cond u)$ to zero and solving for respectively $q(u)$ and $q(z\cond u)$ yields that
\begin{align}
    q(u) &\propto p(u) \int p(z) \exp\la\log p(e\cond f)\ra_{p(f\cond u, z)}\id{z}, \label{eq:qu-smf} \\
    q(z\cond u) &\propto p(z)\exp\la \log p(e\cond f)\ra_{p(f\cond u, z)}, \label{eq:qz-smf}
\end{align}
which are computed in \cref{app:computation_quz}.
As opposed to \cref{eq:qu,eq:qz}, \cref{eq:qu-smf,eq:qz-smf} are uncoupled, but $q(u)$'s moments are now intractable. We can, however, evaluate $q(u)$. We therefore use elliptical slice sampling (ESS) \cite{Murray:2010:Elliptical_Slice_Sampling} to sample from $q(u)$ and use these samples to approximate $q(f, u, z)$. To help mixing the Markov chain, we initialise the sampler with the MF approximation.

\section{Experiments}
We test the CGPCM by applying it to synthetic and real-world signals. We show that in certain situations the causality constraint provides a useful inductive bias that leads to better predictions.

@@

\begin{figure*}[t]
    \centering
    GPCM
    \begin{tabularx}{\linewidth}{
            >{\centering}>{\hsize=.43\hsize}X  % ??
            >{\centering}>{\hsize=.2\hsize}X
            >{\centering}>{\hsize=.2\hsize}X
            >{\centering}>{\hsize=.2\hsize}X
        }
        $f\cond h$ & $h$ & $k_{f\cond h}$ & PSD of $f\cond h$
    \end{tabularx}
    \includegraphics[width=\linewidth, height=3cm]{resources/cropped/learning_known_kernels_acausal_sample_gpcm.pdf}
    CGPCM
    \begin{tabularx}{\linewidth}{
            >{\centering}>{\hsize=.43\hsize}X  % ??
            >{\centering}>{\hsize=.2\hsize}X
            >{\centering}>{\hsize=.2\hsize}X
            >{\centering}>{\hsize=.2\hsize}X
        }
        $f\cond h$ & $h$ & $k_{f\cond h}$ & PSD of $f\cond h$
    \end{tabularx}
    \includegraphics[width=\linewidth, height=3cm]{resources/cropped/learning_known_kernels_acausal_sample_cgpcm.pdf}
    \caption{Predictions of the GPCM and CGPCM after learning a noisy random sample from the GPCM. The sample consists of 300 data points. We used 51 evenly-spaced inducing points for $h$ and 80 evenly-spaced inducing points for $x$; inducing points are indicated by black dots in the figure. All parameters were initialised randomly and the procedure from \cref{sec:inference} was followed. All signals are normalised to unity power. {\color{red}Red} denotes truth; {\color{darkgreen}green} denotes observations, the emperical autocorrelation, and the periodogram for respectively the function $f\cond h$, the kernel $k_{f\cond h}$, and the PSD of $f\cond h$; {\color{blue}blue} denotes quantities learned through the MF approximation; and {\color{mycyan}cyan} denotes quantities learned through the SMF approximation. For any learned quantity it holds that lines correspond to means and areas correpond to 95\% central credible regions.}
    \label{fig:toy_acausal_sample}
\end{figure*}

@@

\begin{figure*}[t]
    \centering
    GPCM
    \begin{tabularx}{\linewidth}{
            >{\centering}>{\hsize=.43\hsize}X  % ??
            >{\centering}>{\hsize=.2\hsize}X
            >{\centering}>{\hsize=.2\hsize}X
            >{\centering}>{\hsize=.2\hsize}X
        }
        $f\cond h$ & $h$ & $k_{f\cond h}$ & PSD of $f\cond h$
    \end{tabularx}
    \includegraphics[width=\linewidth, height=3cm]{resources/cropped/learning_known_kernels_causal_sample_gpcm.pdf}
    CGPCM
    \begin{tabularx}{\linewidth}{
            >{\centering}>{\hsize=.43\hsize}X  % ??
            >{\centering}>{\hsize=.2\hsize}X
            >{\centering}>{\hsize=.2\hsize}X
            >{\centering}>{\hsize=.2\hsize}X
        }
        $f\cond h$ & $h$ & $k_{f\cond h}$ & PSD of $f\cond h$
    \end{tabularx}
    \includegraphics[width=\linewidth, height=3cm]{resources/cropped/learning_known_kernels_causal_sample_cgpcm.pdf}
    \caption{Predictions of the GPCM and CGPCM after learning a noiseless random sample from the CGPCM. The sample consists of 300 data points. We used 51 evenly-spaced inducing points for $h$ and 80 evenly-spaced inducing points for $x$; inducing points are indicated by black dots in the figure. All parameters were initialised randomly and the procedure from \cref{sec:inference} was followed. All signals are normalised to unity power. {\color{red}Red} denotes truth; {\color{darkgreen}green} denotes observations, the emperical autocorrelation, and the periodogram for respectively the function $f\cond h$, the kernel $k_{f\cond h}$, and the PSD of $f\cond h$; {\color{blue}blue} denotes quantities learned through the MF approximation; and {\color{mycyan}cyan} denotes quantities learned through the SMF approximation. For any learned quantity it holds that lines correspond to means and areas correpond to 95\% central credible regions.}
    \label{fig:toy_causal_sample}
\end{figure*}

@@

Finally, note that in nearly every case predictions through the SMF approximation have better calibrated credible regions than predictions through the MF approximation.
% \subsection{Speech}
% Coming up.

@@

\subsection{Head-Related Impulse Response Estimation}
The head-related impulse response (HRIR) describes how the ear's reception of sound depends on the direction from which the sound comes. Specifically, the received sound is convolved with the HRIR before it is sensed by the ear, and the HRIR depends on the direction from which the sound comes. We consider the problem of inferring the HRIR after some sound has been sensed. Suppose that the sound is background noise and can be modelled by white noise. Then assuming a Gaussian process prior over the HRIR recovers the CGPCM exactly; therefore, in that case, we can infer the HRIR via inferring the filter $h$ in the CGPCM.

\begin{figure*}[t]
    \centering
    \begin{tabularx}{\linewidth}{>{\centering}X>{\centering}X}
        GPCM & CGPCM
    \end{tabularx}
    \begin{tabularx}{\linewidth}{>{\centering}X>{\centering}X>{\centering}X>{\centering}X}
        $k_{f\cond h}$ & $h$ & $k_{f\cond h}$ & $h$
    \end{tabularx}
    \includegraphics[width=.49\linewidth, height=3cm]{resources/hrtf_gpcm.pdf}
    \includegraphics[width=.49\linewidth, height=3cm]{resources/hrtf_cgpcm.pdf}
    \caption{Predictions of the GPCM and CGPCM after learning a HRIR convolved with white noise. The sample consists of 500 data points. We used 101 evenly-spaced inducing points for $h$ and 150 evenly-spaced inducing points for $x$; inducing points are indicated by black dots in the figure. All parameters were initialised randomly and the procedure from \cref{sec:inference} was followed. All signals are normalised to unity power. {\color{red}Red} denotes truth, {\color{blue}blue} denotes quantities learned through the MF approximation, and {\color{mycyan}cyan} denotes quantities learned through the SMF approximation. For any learned quantity it holds that lines correspond to means and areas correpond to 95\% central credible regions.}
    \label{fig:hrir}
\end{figure*}

@@

\section{Discussion}
We have presented the CGPCM as a model of causal, spectrally complex dynamical phenomena. Inference in the CGPCM is performed in a two-step procedure: first a collapsed varational MF approach is used to obtain an initial approximation, and then this approximation is refined through a variational SMF approach combined with ESS. The CGPCM has been tested on synthesised and real-world signals and shows encouraging results. In particular, the CGPCM shows the capability of modelling rough signals without attributing this roughness to observation noise.
